HO unification from object language to meta language

### 2024-03-25 paper/main.tex
progress can change the link store

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}}

@@

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{\{l_1,l_2\}}{\rho}
\end{array}
$$

@@

\noindent
TODO: about the cut

@@

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification variable and the
number of abstractions in its assignment have to match.

@@

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems) that play no role in this section
but play a major role in Sections \ref{sec:eta} and \ref{sec:beta}.

@@

XXX links are update unlike section 2
\input{code/hstep}
\input{code/progress}

@@

\subsection{Decompilation}

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}}

@@

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}{\_}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

We can define $s_1 \Uo{} s_2$ by specializing the code of \hrun{} to
$\mathcal{S} = \{ s_1, s_2 \}$ as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress~(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

@@

\noindent
Note that the cut operator is inessential, it could be
removed at the cost of a verbose test on the head of \elpiIn{L}
in the last rule (\elpiIn{L} head
can be \elpiIn{fcon}, \elpiIn{flam} or a name).

Applying the substitution in \Ho{} is very similar, with
the caveat that assignments have to be moved to the
current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.

@@

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification variable and the
number of abstractions in its assignment have to match. In turn
this grants that \elpiIn{move} never fails.

\input{code/move}

@@

The signature of the \elpiIn{comp} predicate below allows for the generation of
links (suspended unification problems) that play no role in this section
but play a major role in Sections \ref{sec:eta} and \ref{sec:beta}.
With respect to \ref{sec:problem-statement} the signature also allows
for updates to the substitution.  The code below only allocates space
for the variables, i.e. sets their memory address to \elpiIn{none},
a details not worth mentioning in the previous discussion.

@@

XXX links are update unlike section 2
\input{code/hstep}

TODOD: discuss signature of progress

\input{code/progress}

@@

TODO: why it terminates

\subsection{Substitution decompilation}
